---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%"
)
```
# admixtools

Lightweight R implementations of some [AdmixTools](https://github.com/DReichLab/AdmixTools) programs.

## Installation

To install `admixtools`, first you need `devtools`:

```{r, eval=FALSE}
install.packages("devtools")
```

Then install `admixtools`, located here on O2:

```{r, eval=FALSE}
pkg = 'file:///n/groups/reich/robert/projects/admixprograms/admixtools_0.1.0.tar.gz'
devtools::install_url(pkg, dependencies='Imports', upgrade='never')
```

Load it like this:

```{r}
library("admixtools")
```


## Usage

If you already have precomputed f2-statistics, you can run estimate admixture weights like this.

```{r}
target = 'Denisova.DG'
left = c('Altai_Neanderthal.DG', 'Vindija.DG')
right = c('Chimp.REF', 'Mbuti.DG', 'Russia_Ust_Ishim.DG', 'Switzerland_Bichon.SG')
```
```{r, eval = FALSE}
f2_dir = '/n/groups/reich/robert/projects/admixprograms/f2blocks_v41.1/'
qpadm(target, left, right, f2_dir = f2_dir)
```

Or you can use these f2-statistics to fit an admixturegraph to the data.

```{r, eval = FALSE}
qpg_results = qpgraph(graph1, f2_dir = f2_dir)
```
```{r, eval = TRUE, echo = FALSE}
qpg_results = qpgraph(graph1, f2_blocks, block_lengths)
```

More examples are in the [`Get started`](articles/admixtools.html) section.


