---
title: "Finding admixture graphs"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
---



## Finding graphs with good fits

Another advantage of being able to quickly evaluate a single model using precomputed f-statistics is that we can evaluate many different graphs in order to find graph topologies with good fits.

<br>


The function `find_graphs` generates and evaluates `numgraphs` admixture graphs in `numgen` iterations across `numrep` independent repeats to find well fitting admixture graphs.

```{r, echo = FALSE}
library(tidyverse)
library(admixtools)
```

```{r, eval = FALSE}
opt_results = find_graphs(my_f2_dir, pops, outpop = pops[1], numrep = 200,
                          numgraphs = 100, numgen = 20, numsel = 5, numadmix = 3)
```
<br>

Once that is done, the following commands will extract the best fitting model overall, and the best fitting model from each independent repeat.

```{r, eval = FALSE}
winner = opt_results %>% top_n(1, -jitter(score))
winners = opt_results %>% group_by(run) %>% top_n(1, -jitter(score)) %>% ungroup
```
```{r, echo = FALSE}
winner = example_winner
winners = example_winners
```

```{r, fig.width = 9, fig.align="center"}
winner$score[[1]]
plot_graph(winner$edges[[1]])
```
<br>


### Similar graphs



### Parallelization

Depending on the input parameters, `find_graphs` can take a long time to run. With the parameters above, 420,000 admixture graphs (200 * 100 * (20+1)) will be generated and evaluated. Thanks to the function [future_map](https://www.rdocumentation.org/packages/furrr/versions/0.1.0/topics/future_map) from the [furrr](https://github.com/DavisVaughan/furrr) package, this can be sped up by parallelizing across the independent repeats. The function [`future::plan`](https://www.rdocumentation.org/packages/future/versions/1.15.1/topics/plan) can be called to specify the details of the parallelization. This can be used to parallelize across CPU cores or across nodes on a compute cluster.

Setting it up like this will make it run multithreaded:
```{r, eval = FALSE}
future::plan('multiprocess')
```
<br>

On the O2 cluster, the following command will set up parallelization across compute nodes.

```{r, eval = FALSE}
future::plan(tweak(batchtools_slurm, workers = 50,
                   resources=list(ncpus = 1, memory = 1024,
                                  walltime = 10*60*60, partition = 'short')))
```
This specifies that up to 50 jobs should be run at a time, with each one requesting one CPU, 1024 MB of memory, and 10 hours on the `short` partition.

This requires the R package `future.batchtools` and a batchtools template file in the working directory, such as this one: `/n/groups/reich/robert/projects/admixprograms/batchtools.slurm.tmpl`.

With this setup the `find_graphs` function will submit each of the 200 repeats as a separate job.

As it will still take a while for this to finish, it is a good idea to submit this as one job which calls an R script. That R script will in turn spawn 200 new jobs and wait for them to finish and return their results.

The R script could look like this.
```{r, eval = FALSE}
library(admixtools)
library(future.batchtools)

future::plan(tweak(batchtools_slurm, workers=50,
                   resources=list(ncpus = 1, memory=1024,
                                  walltime=10*60*60, partition='short')))

pops = c('popA', 'popB', 'popC', 'popD')
opt_results = find_graphs('/my/f2/dir/', pops, outpop = pops[1], numrep = 200,
                          numgen = 20, numgraphs = 100, numadmix = 3, verbose = FALSE)

save(opt_results, file='opt_results.RData')
```

It could be in a file called `opt_graphs.Rscript` and be run like this, or submitted as a job.
```{r, eval = FALSE}
Rscript opt_graphs.Rscript
```

It takes more time to evaluate larger graphs, and in particular graphs with more admixture nodes. It's probably a good idea to start with a small number of repeats, generations, and graphs per generation to get a sense of the runtime before scaling it up.

<br>



## Summarizing graphs

<br>

> *All models are wrong, but some are useful.*   --- George Box

<br>

> *All good models are alike; each bad model is bad in its own way.*   --- Leo Tolstoy

<br>

For any set of populations there may be many graphs which can explain the data about equally well. Rather than declaring one of these models as the correct model, it may be more useful to find features that are shared between all models with good fits. The following function attempts to find such features by summarizing a set of graphs. It counts how often each population triple occurs in which configuration across all graphs.

```{r, eval = FALSE}
triples = summarize_triples(winners)
```
```{r, echo = FALSE}
triples = example_triples
```

```{r}
triples %>% arrange(-clade) %>% slice(1)
```

The output for this population triple (1: `Rus`, 2: `Swi`, 3: `Den`) can be read like this:

* `numgraphs`: 200 graphs were compared
* `clade`: In 34.5% of all graphs `Rus` and `Swi` form a clade with respect to `Den`
* `x13`: In 0.5% of all graphs `Rus` is closer to `Den` than `Swi` is to `Den`, with ancestors of `Rus` admixing into `Den`
* `x31`: In 50.5% of all graphs `Rus` is closer to `Den` than `Swi` is to `Den`, with ancestors of `Den` admixing into `Rus`
* `x23`: In 24.5% of all graphs `Swi` is closer to `Den` than `Rus` is to `Den`, with ancestors of `Swi` admixing into `Den`
* `x32`: In 38.0% of all graphs `Swi` is closer to `Den` than `Rus` is to `Den`, with ancestors of `Den` admixing into `Swi`
* `toptopo`: The most common topology for this triple across all graphs. `011111` specifies a topology where condition `x13` is not satisfied, but conditions `x23`, `x31`, `x32`, `x12`, and `x21` are satisfied. A topology starting with `0000` is one where the first two populations form a clade.
* `toptopocnt`: The number of times `toptopo` was observed
* `topos`: The number of times each topology was observed


<br>



## Finding valid qpAdm models

Every admixture graph maps to a specific set of valid qpAdm models. The function `qpadm_models` lists all valid qpAdm models for a graph. Models which are contained within larger valid models are not shown. The number of valid qpAdm models can be very large for big graphs, so you should only run this on small graphs.

```{r, fig.width = 9, fig.align="center"}
mygraph = winners %>% slice(6) %$% igraph %>% pluck(1)
plot_graph(mygraph)
qpadm_models(mygraph, add_outgroup = TRUE, nested = FALSE, abbr = 3)
```

The only qpAdm models which are valid under this graph have `Den` as target and `Chi`, `Rus`, and `Swi` as right populations.


Valid qpAdm models need to satisfy the following criteria:

* There have to be more right populations than left populations
* The left populations should not form a clade with respect to any right population. This means that a qpgraph outgroup is not an informative right population, but it can be included anyway, with `add_outgroup`.
* Each set of left populations which forms a source for the target population has to form a clade with target at the exclusion of each right population. (Note: Check if that's correct and if it can be stated in a better way.)







