---
title: "Genotype data"
author: "Robert Maier"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Genotype data}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

**Under construction**

ADMIXTOOLS 2 includes functions for reading genotype data that may be useful outside the scope of demographic inference.

While there are other R packages for reading and writing binary PLINK files ([plink2R](https://github.com/gabraham/plink2R), [genio](https://github.com/OchoaLab/genio)), no such packages exist for binary EIGENSTRAT/PACKEDANCESTRYMAP format files yet exist. Some of the functions shown here can also read only a subset of all SNPs (as long as they are consecutive) or samples.

### Reading genotype files

```{r, eval = FALSE}
geno = read_plink("plink/prefix")
geno = read_packedancestrymap("packedancestrymap/prefix")
```


### Computing allele frequencies

```{r, eval = FALSE}
afs = plink_to_aftable("plink/prefix")
afs = packedancestrymap_to_aftable("packedancestrymap/prefix")
```


### Converting PACKEDANCESTRYMAP to PLINK

This function's original purpose was to extract samples from genotype files, but it can also be used to convert genotype files from PACKEDANCESTRYMAP to PLINK format: If no samples are provided, it defaults to all samples.

ADMIXTOOLS 2 currently doesn't provide the option to write binary PACKEDANCESTRYMAP files, so the conversion only goes one way.

```{r, eval = FALSE}
extract_samples("packedancestrymap/prefix", "plink_output")
```





