---
title: "Genotype data"
author: "Robert Maier"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Genotype data}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

**Under construction**

*ADMIXTOOLS 2* can read genotype data in binary *PLINK* format which is described [here](https://www.cog-genomics.org/plink2/input), and in *EIGENSTRAT*/*PACKEDANCESTRYMAP* formats, described [here](https://reich.hms.harvard.edu/software/InputFileFormats). 

While there are other R packages for reading and writing binary *PLINK* files ([plink2R](https://github.com/gabraham/plink2R), [genio](https://github.com/OchoaLab/genio)), no such packages exist to date for *PACKEDANCESTRYMAP* format files, so some of the functions shown here may be useful outside of the *ADMIXTOOLS* framework.

Note that the functions below impose the restriction than individual IDs may not be duplicated across populations or families.


*ADMIXTOOLS 2* currently doesn't handle many common genotype data wrangling tasks such merging different data sets. That's partly because *PLINK* already excels at this task, and partly because it requires a large amount of effort to handle all possible scenarios that can come up, without introducing potentially dangerous bugs. If converting your data to *PLINK* format, and using *PLINK* for these tasks is not a practical solution in your case, please let me know about it.


### Reading genotype files

```{r, eval = FALSE}
geno = read_plink("plink/prefix")
geno = read_eigenstrat("eigenstrat/prefix")
geno = read_packedancestrymap("packedancestrymap/prefix")

# read SNPs 1000 to 2000
geno = read_packedancestrymap("packedancestrymap/prefix", first = 1000, last = 2000)
# read only ind1 and ind2
geno = read_packedancestrymap("packedancestrymap/prefix", inds = c('ind1', 'ind2'))
```


### Computing allele frequencies

```{r, eval = FALSE}
afs = plink_to_afs("plink/prefix")
afs = eigenstrat_to_afs("eigenstrat/prefix")
afs = packedancestrymap_to_afs("packedancestrymap/prefix")
```


### Converting *PACKEDANCESTRYMAP* to *PLINK*

This function converts genotype files from *PACKEDANCESTRYMAP* or *EIGENSTRAT* to *PLINK* format. If `inds` or `pops` is specified, only those samples will be extracted.

*ADMIXTOOLS 2* currently doesn't provide the option to write binary *PACKEDANCESTRYMAP* files, so the conversion only goes one way.

```{r, eval = FALSE}
eigenstrat_to_plink("eigenstrat_input_prefix", "plink_output_prefix")
packedancestrymap_to_plink("packedancestrymap_input_prefix", "plink_output_prefix")
```





