---
title: "Data formats"
author: "Robert Maier"
date: "2022-11-12"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Data formats}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---


## Genotype data formats 

*ADMIXTOOLS 2* can read genotype data in four formats.

1. Binary *PLINK* format (PACKEDPED), described [here](https://www.cog-genomics.org/plink2/input)
2. Binary *PLINK 2* format (PFILE), described [here](https://www.cog-genomics.org/plink/2.0/input)
3. Binary *PACKEDANCESTRYMAP* format
4. Text based *EIGENSTRAT* format

*PACKEDANCESTRYMAP* and *EIGENSTRAT* are described [here](https://reich.hms.harvard.edu/software/InputFileFormats)

In all four formats a dataset consists of three files: one file for the genotype matrix, one file for the SNP metadata, and one file for the sample metadata.

The following shows what a data set for three SNPs and five samples grouped into two populations (French.DG, Onge.DG) could look like:

### PLINK

PLINK `.fam` files, have six columns, but only the first two (family/population ID and sample ID) are relevant in *ADMIXTOOLS 2*

```
French.DG	S_French-1.DG	0	0	M	-9
French.DG	S_French-2.DG	0	0	F	-9
French.DG	B_French-3.DG	0	0	M	-9
Onge.DG	BR_Onge-2.DG	0	0	F	-9
Onge.DG	BR_Onge-1.DG	0	0	F	-9
```

PLINK `.bim` file:

```
1	rs3094315	0.02013	752566	G	A
1	rs12124819	0.020242	776546	A	G
1	rs28765502	0.022137	832918	T	C
```

### PLINK 2 (PFILE)

A *PLINK 2* dataset consists of three files: `.pgen` (binary genotype matrix), `.pvar` (variant metadata), and `.psam` (sample metadata). The prefix convention is the same as for binary *PLINK* (one prefix that resolves to all three), so *ADMIXTOOLS 2* auto-detects PFILE format whenever `<prefix>.pgen`, `<prefix>.pvar`, and `<prefix>.psam` all exist at the prefix you pass.

`.psam` is a tab-separated text file. Population labels live in the `#FID` column when present (matching the PLINK convention of FID-as-population); when `#FID` is missing or `0`, *ADMIXTOOLS 2* falls back to `IID`:

```
#FID	IID	SEX
French.DG	S_French-1.DG	1
French.DG	S_French-2.DG	2
French.DG	B_French-3.DG	1
Onge.DG	BR_Onge-2.DG	2
Onge.DG	BR_Onge-1.DG	2
```

`.pvar` is tab-separated text and (unlike `.bim`) supports multi-allelic sites by allowing comma-separated `ALT` alleles. The `pfile_to_afs()` function accepts a `multiallelic = c("error", "skip", "first_alt")` argument to control how those sites are handled (default `"error"`):

```
#CHROM	POS	ID	REF	ALT
1	752566	rs3094315	G	A
1	776546	rs12124819	A	G
1	832918	rs28765502	T	C
```

`.pvar` may also be supplied in Zstd-compressed form as `.pvar.zst` — *ADMIXTOOLS 2* picks up either form transparently. Reading `.pvar.zst` requires the `zstd` CLI on PATH (the genotype access via `pgenlibr` handles `.zst` natively, but the R-side metadata parser shells out to `zstd -d`).

Note: `.pvar` does *not* by default carry centimorgan positions. When the `CM` column is absent *ADMIXTOOLS 2* leaves `cm = 0` for all variants and falls back to base-pair distance for block partitioning (`blgsize = 1e6` works well at typical block sizes). Alternatively you can supply an external recombination map via `cm_file =` (a TSV with a SNP-identifier column — `variant_id`, `SNP`, or `ID` — and a `cm` column).

### EIGENSTRAT/PACKEDANCESTRYMAP

*EIGENSTRAT* and *PACKEDANCESTRYMAP* use the same format for the metadata. The only difference is that the genotype data is stored in a text file in *EIGENSTRAT*, but in a binary file in *PACKEDANCESTRYMAP*.

The `.ind` file has 3 out of the 6 columns in the `.fam` file, in different order:

```
S_French-1.DG M French.DG
S_French-2.DG F French.DG
B_French-3.DG M French.DG
BR_Onge-2.DG F Onge.DG
BR_Onge-1.DG F Onge.DG
```

`.snp` files are the same as `.bim` files, except that the first two columns are swapped:

```
rs3094315 1	0.02013	752566	G	A
rs12124819 1	0.020242	776546	A	G
rs28765502 1	0.022137	832918	T	C
```

*EIGENSTRAT* `.geno` files look like this:

```
29101
10211
22901
```

One SNP per row, one sample per column, and missing data is denoted by `9`. 


## Reading genotype files

While there are other R packages for reading and writing binary *PLINK* files ([plink2R](https://github.com/gabraham/plink2R), [genio](https://github.com/OchoaLab/genio)), no such packages exist to date for *PACKEDANCESTRYMAP* format files, so some of the functions shown here may be useful outside of the *ADMIXTOOLS* framework.

Note that the functions below impose the restriction than individual IDs may not be duplicated across populations or families.


```{r, echo = FALSE, warning = FALSE, message = FALSE}
library(magrittr)
library(tidyverse)
library(admixtools)
```


```{r, eval = FALSE}
geno = read_plink("plink/prefix")
geno = read_eigenstrat("eigenstrat/prefix")
geno = read_packedancestrymap("packedancestrymap/prefix")

# read SNPs 1000 to 2000
geno = read_packedancestrymap("packedancestrymap/prefix", first = 1000, last = 2000)
# read only ind1 and ind2
geno = read_packedancestrymap("packedancestrymap/prefix", inds = c('ind1', 'ind2'))
```

For PFILE, there is no `read_pfile` — PFILE inputs are consumed directly by the f-statistic pipeline (`extract_f2`, `f2_from_geno`, `f4blockdat_from_geno`, `qpfstats`) which auto-detect the format from the file extensions at the prefix.


## Extracting populations

Populations can be extracted from *PLINK* files (and written to another set of *PLINK* files) like this: 

```{r, eval = FALSE}
extract_samples("input/prefix", "output/prefix", pops = c("pop1", "pop2"))
```

`extract_samples` can also be used to extract samples, similar to `plink --keep`.


## Combining data sets

*ADMIXTOOLS 2* can't merge data sets. Data in *PLINK* format can be merged using `plink --bmerge`. 


## Computing allele frequencies

```{r, eval = FALSE}
afs = plink_to_afs("plink/prefix")
afs = eigenstrat_to_afs("eigenstrat/prefix")
afs = packedancestrymap_to_afs("packedancestrymap/prefix")
afs = pfile_to_afs("plink2/prefix")

# pfile_to_afs supports the same range/subset args as the others
afs = pfile_to_afs("plink2/prefix", inds = c('ind1', 'ind2'))
afs = pfile_to_afs("plink2/prefix", pops = c('pop1', 'pop2'))
afs = pfile_to_afs("plink2/prefix", first = 1000, last = 2000)

# When .pvar lacks centimorgan positions (the default), supply a map
# via cm_file (TSV with variant_id + cm columns)
afs = pfile_to_afs("plink2/prefix", cm_file = "ref_genmap.tsv")
```


## Converting *PACKEDANCESTRYMAP* to *PLINK*

This function converts genotype files from *PACKEDANCESTRYMAP* or *EIGENSTRAT* to *PLINK* format. If `inds` or `pops` is specified, only those samples will be extracted.

*ADMIXTOOLS 2* currently doesn't provide the option to write binary *PACKEDANCESTRYMAP* files, so the conversion only goes one way.

```{r, eval = FALSE}
eigenstrat_to_plink("eigenstrat_input_prefix", "plink_output_prefix")
packedancestrymap_to_plink("packedancestrymap_input_prefix", "plink_output_prefix")
```


## Admixture graph formats

There are several ways in which admixture graphs can be represented. Some of them are useful because they can be stored as human readable text files, and others are useful because they make it easier for R to process graphs. *ADMIXTOOLS 2* has several function for importing, exporting and converting between formats.

Before fitting an admixture graph, the graph only has nodes and edges. After fitting a graph, each edge has a weight associated with it, and maybe some other information. Here I mostly show examples of graphs that are not fitted yet.

### Edge list format

The simplest way to represent a graph in a text file is by writing each edge in the graph in a separate line:

```
R      O
R      n1
n1     n2
n1     n3
n2     A
n2     admix
n3     B
n3     n4
n4     C
n4     admix
admix  D
```

We can save this in a text file `graph.txt` and read it into R like this:

```{r, eval = FALSE}
graph_el = read_table('~/Downloads/graph.tsv', col_names = F)
plotly_graph(graph_el)
```
```{r, echo = FALSE}
graph_el = tribble(~from, ~to,
'R', 'O',
'R', 'n1',
'n1', 'n2',
'n1', 'n3',
'n2', 'A',
'n2', 'admix',
'n3', 'B',
'n3', 'n4',
'n4', 'C',
'n4', 'admix',
'admix', 'D')
plotly_graph(graph_el)
```


`qpgraph()` uses this format to represent fitted graphs. In addition to the first two columns (`from` and `to`), there will be additional columns to indicate the type of edge, its estimated weight, and possibly more information.

```{r}
qpgraph(example_f2_blocks, example_igraph)$edges %>% head
```


### igraph format

Many *ADMIXTOOLS 2* functions use the `igraph` package, which has its own graph format. The most important *ADMIXTOOLS 2* functions work with graphs in either the edge list or the `igraph` format, but some only work with graphs in the `igraph` format. It's easy to convert between the two formats:


```{r}
class(graph_el)
graph = edges_to_igraph(graph_el)
class(graph)
graph_el2 = as.data.frame(igraph::as_edgelist(graph))
class(graph_el2)
```


### Original ADMIXTOOLS format

The text representation of the graph above in the original *ADMIXTOOLS* software would look like this:

```
vertex              R
vertex              O
vertex              n1
vertex              n2
vertex              n3
vertex              A
vertex              admix
vertex              B
vertex              n4
vertex              C
vertex              D
label               O                   O
label               A                   A
label               B                   B
label               C                   C
label               D                   D
edge                R_O                 R                   O
edge                R_n1                R                   n1
edge                n1_n2               n1                  n2
edge                n1_n3               n1                  n3
edge                n2_A                n2                  A
edge                n3_B                n3                  B
edge                n3_n4               n3                  n4
edge                n4_C                n4                  C
edge                admix_D             admix               D
admix               admix               n2                  n4
```

If this is stored in a file called `graph.txt`, it can be imported into R like this:

```{r, eval = FALSE}
graph_el = parse_qpgraph_graphfile('graph.txt')
```

### DOT format

DOT format is often used for plotting admixture graphs. You can export a fitted admixture graph like this:

```{r}
fit = qpgraph(example_f2_blocks, example_igraph)$edges
write_dot(fit, 'graph.dot')
```

You can read a DOT file back into an igraph with `parse_dot()`:

```{r, eval = FALSE}
graph = parse_dot("graph.dot")
```

`parse_dot()` is a minimal reader that recovers the topology from `from -> to` edges. Edge attributes (weights, types) are not preserved on the round trip; for full round-tripping of fitted graphs, use the edge-list format instead.

