---
title: "Genotype data"
author: "Robert Maier"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Genotype data}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

**Under construction**

*ADMIXTOOLS 2* includes functions for reading genotype data that may be useful outside the scope of demographic inference.

While there are other R packages for reading and writing binary *PLINK* files ([plink2R](https://github.com/gabraham/plink2R), [genio](https://github.com/OchoaLab/genio)), no such packages exist to date for binary *EIGENSTRAT*/*PACKEDANCESTRYMAP* format files. Some of the functions shown here can also read only a subset of all samples or SNPs (as long as they are consecutive).

### Reading genotype files

```{r, eval = FALSE}
geno = read_plink("plink/prefix")
geno = read_packedancestrymap("packedancestrymap/prefix")
# read SNPs 1000 to 2000
geno = read_packedancestrymap("packedancestrymap/prefix",
                              first = 1000, last = 2000)
# read only ind1 and ind2
geno = read_packedancestrymap("packedancestrymap/prefix",
                              inds = c('ind1', 'ind2'))
```


### Computing allele frequencies

```{r, eval = FALSE}
afs = plink_to_aftable("plink/prefix")
afs = packedancestrymap_to_aftable("packedancestrymap/prefix")
```


### Converting *PACKEDANCESTRYMAP* to *PLINK*

This function converts genotype files from *PACKEDANCESTRYMAP* to *PLINK* format. If `inds` or `pops` is specified, only those samples will be extracted.

*ADMIXTOOLS 2* currently doesn't provide the option to write binary *PACKEDANCESTRYMAP* files, so the conversion only goes one way.

```{r, eval = FALSE}
packedancestrymap_to_plink("packedancestrymap_input_prefix", "plink_output_prefix")
```





