---
title: "Round-tripping admixture graphs through LEGOFIT"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Round-tripping admixture graphs through LEGOFIT}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
# Cached LEGOFIT artifacts ship under inst/extdata/legofit and are read with
# system.file() so the vignette builds without the external binary.
ext <- function(f) system.file("extdata/legofit", f, package = "admixtools")
library(admixtools)
```

## Why LEGOFIT

[LEGOFIT](https://github.com/alanrogers/legofit) estimates population history
from the frequencies of nucleotide site patterns. It takes a description of an
admixture graph (a `.lgo` file) and a file of observed site-pattern
frequencies, then fits the free parameters of the graph (split times,
population sizes, admixture fractions) to the data.

ADMIXTOOLS 2 can now hand graphs to LEGOFIT and read fitted results back. This
article walks the full loop: an ADMIXTOOLS edge tibble is written out to a
`.lgo` file, LEGOFIT fits it, and the fitted parameters come back as an edge
tibble you can keep working with. `graph_to_lgo()` provides the write
direction and `read_legofit_output()` the read direction, with `run_legofit()`
as a thin convenience wrapper over the external binary.

LEGOFIT itself is an external command-line tool, not an R dependency. The
chunks below that would invoke it are shown but not run; their results are
loaded from small precomputed files that ship with the package, so this
vignette builds anywhere.

## Write: a graph to a .lgo file

We start from a small three-leaf tree with no admixture. The leaves are `x`,
`y`, and `z`; `xy` is their internal ancestor and `xyz` is the root. The
`time` column gives node ages, oldest at the root.

```{r demo-graph}
demo_graph <- tibble::tribble(
  ~from,  ~to,   ~type,    ~weight,  ~time,
  "xyz",  "xy",  "normal", NA_real_, 1.5,
  "xyz",  "z",   "normal", NA_real_, 2,
  "xy",   "x",   "normal", NA_real_, 0.5,
  "xy",   "y",   "normal", NA_real_, 0.5
)
demo_graph
```

`graph_to_lgo()` turns the tibble into LEGOFIT's `.lgo` grammar: a block of
time and population-size parameter declarations, one `segment` per node, and
the `derive` relationships that wire the topology together.

```{r write-lgo}
cat(graph_to_lgo(demo_graph), sep = "\n")
```

Split times that are free to be fitted appear as `time free`, fixed leaf times
as `time fixed`. Each leaf segment carries `samples=1`.

## Fit: run LEGOFIT

With a `.lgo` file and a site-pattern file you call LEGOFIT. `run_legofit()`
wraps the external binary: it writes the graph to a temporary `.lgo`, invokes
`legofit`, and parses the fitted output back through `read_legofit_output()`.
The site-pattern file used here (`demo.opf`) is LEGOFIT's bundled `ex1`
example, vendored alongside the cached output.

The next chunk is not evaluated when the vignette builds, because it needs the
external `legofit` binary. It is the exact call that produced the cached
result shown below.

```{r fit, eval = FALSE}
fit <- run_legofit(
  demo_graph,
  patterns = ext("demo.opf"),
  args     = c("--threads", "1", "-1", "-d", "0")
)
```

Instead of running it, we load the cached fit:

```{r fit-cached}
fit <- readRDS(ext("demo_fit.rds"))
fit
```

## Read: the fitted edge tibble

`read_legofit_output()` returns the input edges with fitted values filled in.
The `time` column now holds the LEGOFIT-fitted age of each edge's younger
endpoint; `twoN` holds the fitted effective population size. For an
admixture graph, `admix_prop` and `admix_event_time` would carry the fitted
mixture fractions and admixture-event times (there is no admixture in this
demo, so those are `NA`).

Two pieces of fit metadata ride along as attributes. `node_times` is the full
set of fitted node ages, including the root, which has no incoming edge and so
does not appear in the `time` column:

```{r node-times}
attr(fit, "node_times")
```

`fit_convergence` records how the optimizer terminated. `status` is
`"reached_goal"` when LEGOFIT converged, along with the final cost and the
swarm spread:

```{r convergence}
attr(fit, "fit_convergence")
```

A third attribute, `identifiability`, flags which fitted values the
site-pattern data can actually constrain (for example, admixture-event times
are structurally non-identifiable). See `?read_legofit_output` for the full
classification.

## Round trip: reading a real .lgo file

The write direction is only half of the interoperability story. LEGOFIT
`.lgo` files written elsewhere can be read straight into an ADMIXTOOLS edge
tibble with `read_lgo()`. Here we read a 20-population graph that ships as a
cached fixture.

```{r read-lgo}
rha20 <- read_lgo(ext("rha20.lgo"))
nrow(rha20)
head(rha20)
```

`read_lgo()` is pure R and needs no external tool, so this chunk runs at build
time. The file parses into its full topology, including internal nodes that
are themselves sampled. LEGOFIT allows a `segment` to be both an internal
ancestor and a sampled population.

When a downstream analysis needs every sampled population to sit on its own
tip, `add_sampled_tips()` splits each sampled internal node by attaching a new
leaf for the sample, leaving the internal node purely structural. Here it adds
the tips for this graph's sampled internal nodes:

```{r add-tips}
nrow(rha20)
nrow(add_sampled_tips(rha20))
```

This is the bridge to `qpgraph()` and the rest of the ADMIXTOOLS graph
machinery, which expect sampled populations at the tips.

## Availability

LEGOFIT is an external tool and is not installed with ADMIXTOOLS 2. Install it
from <https://github.com/alanrogers/legofit> and put the `legofit` binary on
your `PATH`, or pass its location to `run_legofit(bin = ...)`. When the binary
cannot be found, `run_legofit()` fails with a clear, typed condition
(`legofit_binary_not_found`) rather than a raw system error, so calling code
can detect and handle the missing tool.
