---
title: "Plotting"
author: "Robert Maier"
date: "2026-05-21"
output:
  rmarkdown::html_vignette:
    toc: false
vignette: >
  %\VignetteIndexEntry{Plotting}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

<style>
  #tocnav {
    display: none;
  }
</style>
  

```{r, include=FALSE}
library(admixtools)
```


## Plot admixture graph

The simplest way to visualize an admixture graph is `plot_graph()`, which returns a static `ggplot`:

```{r}
plot_graph(example_igraph)
```

For an interactive 3D version that's easier to inspect inner-node labels on, use `plotly_graph()`:

```{r, warning = FALSE}
plotly_graph(example_igraph)
```

For an external Graphviz `dot`-rendered version (useful for papers, presentations, or batch rendering pipelines), use `write_dot()`. It supports several display options:

```{r, eval = FALSE}
# Basic export
write_dot(example_igraph, "graph.dot")

# Color scheme + label tweaks
write_dot(example_igraph, "graph.dot",
          color = FALSE,            # B&W
          fontsize = 18,            # larger labels
          hide_weights = TRUE)      # drop the drift-value annotations

# Highlight edges that are not identifiable from the data
write_dot(example_igraph, "graph.dot",
          highlight_unidentifiable = TRUE)
```

See the [data formats vignette](io.html) for the full set of `write_dot()` args and round-tripping with `parse_dot()`.


## Admixture graph on a map

`plot_graph_map()` overlays the graph on a geographic map, using per-individual coordinates (e.g., from `example_anno`) to place leaf nodes. It returns an interactive 3D plotly widget - best explored interactively rather than as a static screenshot:

```{r, eval = FALSE}
plot_graph_map(example_igraph, example_anno)
```


## All samples on a map

<div class="bla">
<style>
.bla {
    margin-left: -20%;
    margin-right: -20%;
}
</style>
```{r, fig.height = 8, out.width="120%", fig.align='center'}
plot_map(example_anno, map_layout = 2)
```
</div>










