---
title: "qpWave and qpAdm"
author: "Robert Maier"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{qpWave and qpAdm}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---


**Under construction**

```{r}
library(magrittr)
library(tidyverse)
library(admixtools)
```

## qpWave

### qpWave and f4

$f_4$-statistics can be used to test whether two pairs of populations form a clade with respect to one another: A significant deviation from zero is evidence against cladality. *qpWave* generalizes this test from two groups with two populations each (four in total) to two groups with any number of populations. In the special case of applying *qpWave* to only four populations, we get the same p-value that we would get for an $f_4$-statistic of those four populations:

```{r}
f4(example_f2_blocks, 'Altai_Neanderthal.DG', 'Vindija.DG', 'Chimp.REF', 'Mbuti.DG')

qpwave(example_f2_blocks, c('Altai_Neanderthal.DG', 'Vindija.DG'), c('Chimp.REF', 'Mbuti.DG'))$rankdrop
```

### Estimating the number of independent gene flows

Testing whether two pairs of populations form clades with each other is the same as asking whether there is only one gene flow event which separates the two pairs ($f_4 = 0$) or whether there must have been more than a single gene flow event between the two pairs ($f_4 != 0$).

In the graph on the left, $L1$ and $L2$ form a clade relative to $R1$ and $R2$, which means $L$ and $R$ are only connected through one gene flow event. On the right they do not form a clade, and they are connected through two independent gene flow events (one going through the root, and the other going through the edge `n1b|n2b`).

:::::: {style="display: flex"}

::: {style="width: 50%"}

```{r, echo = F, out.width="100%"}
matrix(c('R', 'n1', 'R', 'n2', 'n1', 'L1',
         'n1', 'L2', 'n2', 'R1', 'n2', 'R2'), , 2, byrow = T) %>%
  plotly_graph
```

:::

::: {style="width: 50%"}

```{r, echo = F, out.width="100%"}
matrix(c('R', 'n1', 'R', 'n2', 'n1', 'n1b', 'n1b', 'L2', 'n1b', 'n2b',
         'n1', 'L1', 'n2b', 'R1', 'n2', 'R2', 'n2', 'n2b'), , 2, byrow = T) %>%
  plotly_graph
```

:::

::::

*qpWave* allows us to estimate a lower bound on the number of gene flow events that separate two groups of populations $L$ and $R$ of size $n_L$ and $n_R$.

The basic idea behind it is that we can use $f_4$-statistics of the form $f_4(L_i, L_j; R_m, R_n)$ to estimate the genetic drift that is shared between pairs of left populations and pairs of right populations. There are $\frac{n_L(n_L-1)}{2} \frac{n_R(n_R-1)}{2}$ $f_4$-statistics of that form, but not all of them contain unique information. In the case where $L$ and $R$ form two clades, all $f_4$-statistics $f_4(L_i, L_j; R_m, R_n)$ are expected to be zero. If there are two independent gene flow events, a few of these $f_4$-statistics will be enough to determine what the other $f_4$-statistics should be. More generally, we can construct a matrix $X$ of these $f_4$-statistics, and the rank of $X$ tells us what the minimum number of gene flows is between $L$ and $R$.

In practice, $X$ is constructed by fixing one population in $L$ and one population in $R$ which results in a matrix of dimensions $(n_L - 1) \times (n_R - 1)$ with all c of the form $f_4(L_1, L_i; R_1, R_j)$. By convention, $n_R > n_L$, which means that the highest possible rank of $X$ is $n_L - 1$. A rank of $n_L - 1$ suggests that at least $n_L - 1$ independent gene flows have occurred between $L$ and $R$.


### Estimating the rank of X

When estimating the rank of $X$, we have to take into account that the $f_4$-statistics are not estimated perfectly. If we didn't take that into account, $X$ would always be of full rank.

To estimate the rank of $X$ while taking into account the noise in the estimates of $f_4$, we test how well $X$ can be approximated by different low rank approximations. An approximation of $X$ with one rank less than full rank can be written as $AB$, where $A$ is $(n_L-1) \times (n_L-2)$, and $B$ is $(n_L-2) \times (n_R-1)$ and $AB$ should be as close as possible to $X$. In this case (one rank less than full rank), $X$ would have rank $n_L - 1$, and $AB$ would have rank $n_L - 2$. If this approximation works well, the residuals $E = X - AB$ will be very small. The tricky part here is that we need to take into account that the $f_4$-statistics are estimated with some amount noise, and that they are highly correlated to each other. To account for that, we 

The difference of these two log likelihoods approximately follows a $$ \Chi^2 $$ distribution with one degree of freedom,

To formally test how much information is added by including T in L, $$ X' $$ is compared to a low rank approximation of itself in which the rank has an upper bound of $$ n_L $$. This low rank approximation of $$ X' $$ can be written as $$ AB $$, where $$ A $$ is $$ n_L \times n_L $$, and $$ B $$ is $$ n_L \times n_R $$ and should minimize the log likelihood $$ L(A,B) = -\frac{1}{2} E'Q^{-1}E  $$. Here, $$ E $$ is a $$ n_L (n_R-1) $$ vector of residuals given by $$ E = X' - AB $$, and $$ Q $$ is the $$ n_L (n_R-1) \times n_L (n_R-1) $$ covariance matrix of all pairs of f4-statistics in $$ X' $$, estimated using block jackknife. An initial estimate of $$ AB $$ is obtained through a singular value decomposition (SVD) of $$ X' $$. If the errors in the estimated f4-statistics were all uncorrelated, this singular value decomposition (SVD) of $$ X' $$ would already yield $$ A $$ and $$ B $$ which minimizes $$ L(A, B) $$. However, since this is generally not the case, $$ A $$ and $$ B $$ are further optimized in an iterative procedure where one matrix is held constant, while the other matrix is adjusted to minimize $$ L(A, B) $$. is compared to a low rank approximation of itself in which the rank has an upper bound of . This low rank approximation of  can be written as , where  is , and  is  and should minimize the log likelihood . Here,  is a  vector of residuals given by , and  is the  covariance matrix of all pairs of f4-statistics in , estimated using block jackknife. An initial estimate of  is obtained through a singular value decomposition (SVD) of . If the errors in the estimated f4-statistics were all uncorrelated, this singular value decomposition (SVD) of  would already yield  and  which minimizes . However, since this is generally not the case,  and  are further optimized in an iterative procedure where one matrix is held constant, while the other matrix is adjusted to minimize .
This likelihood is then compared to the likelihood of the full model, where $$ X' $$ is decomposed into two matrices without constraining the rank to be $$ n_L $$. The difference of these two log likelihoods approximately follows a $$ \Chi^2 $$ distribution with one degree of freedom, and this makes it possible to test whether the populations in L fully explain the variation in T, relative to R. The more general case where rank differences can be greater than one is described in [cite haak]. is decomposed into two matrices without constraining the rank to be . The difference of these two log likelihoods approximately follows a  distribution with one degree of freedom, and this makes it possible to test whether the populations in L fully explain the variation in T, relative to R. The more general case where rank differences can be greater than one is described in [cite haak].
The admixture weights for the populations in L relative to T are given by left null space of the matrix $$ A $$, normalized so that they sum to one., normalized so that they sum to one.


## qpAdm

### Modelling an admixed population

*qpAdm* extends *qpWave* by adding an additional target population ($T$) to the left and right population sets $L$ and $R$, in order to assess whether $T$ can be modeled as a mixture of populations in $L$, at the exclusion of gene flow from populations in $R$.
The null hypothesis tested by *qpAdm* is that the populations in $L$ can account for all shared drift between $T$ and $R$. A significant p-value rejecting this null hypothesis indicates that there has been gene flow between $T$ and $R$ (in either direction) that is not accounted for by any population in $L$.

Often the goal of *qpAdm* modeling is to identify a set of populations $L$ for which this null hypothesis does not fail (i.e. there was no gene flow between $R$ and $T$, relative to $L$), and to estimate the relative contribution of the populations in $L$ to $T$. If no set of populations $L$ can be found which fully explain the variation in $T$ relative to $R$ (all models fail), this may hint at the existence of a yet undiscovered “ghost” population which is related to a population in $R$, but not represented by any of the populations in $L$.
If a set of populations $L$ can be found which fully explains the variation in $T$ relative to $R$, $T$ can be thought of as a mixture of source populations which are represented by populations in $L$, and *qpAdm* will estimate the mixture proportions. However, it is important to keep in mind that the true source populations are almost certainly not identical to any populations in $L$, and that any inferences about the mixture proportions of the true, unobserved sources depends on an unverifiable assumption: Each left population needs to be strictly cladal with one of the true, unobserved source populations.

### How it works

Briefly, *qpAdm* is based on comparing to matrices of $f_4$-statistics: The first matrix captures shared drift between populations in $L$ and in $R$, and the second matrix captures shared drift between populations in $L \cup T$ and populations in $R$. If both matrices have full rank, it suggests that there was gene flow between $T$ and $R$ which does not pass through any populations in $L$. In other words, $T$ cannot be modeled as a mix of populations in $L$, without missing a component that is only shared between $T$ and $R$.

The same in more words:

The starting point of *qpAdm* is the identification of sets of populations $L$ and $R$ such that no pair of populations in $L$ forms a clade relative to all populations in $R$. If that condition is met, the matrix of $f_4$-statistics ($X$) will have full rank. $T$ is then added to the populations in $L$, which creates a new matrix of $f_4$-statistics, $X'$ of dimensions $n_L \times (n_R - 1)$. If the rank of $X'$ is significantly higher than the rank of $X$, this indicates that the new f4-statistics contain information about shared drift which was not already present in the old $f_4$-statistics, and that gene flow has occurred between $R$ and $T$ which is not captured by any population in $L$. On the other hand, if the rank of $X'$ does not increase significantly, it suggests that $T$ does not share any drift with $R$ which is not captured by $L$ (there are no paths from $T$ to any population in $R$ which do not pass through ancestors of a population in $L$). This implies that all shared drift between $T$ and $R$ can be modeled as a linear combination of shared drift between $L$ and $R$.

#### Estimating admixture proportions

If the condition mentioned above is met ($f_4$-statistics involving $T$ can be modeled as a linear combination of $f_4$-statistics without $T$), then the coefficients of this linear combination correspond to the relative contributions of the source populations.


#### Testing rank differences




The weights of the linear combination of $f_4$-statistics excluding $T$ which can correspond to the relative contributions of the source populations. The weights $w$ are normalized so that they sum to one, and they are equivalent to the left null space of the matrix $A$ (they satisfy $w A = 0$), where $A B$ is a rank $n_L - 1$ approximation of $X'$. will have full rank. T is then added to the populations in L, forming a new matrix of f4-statistics,  of dimensions . If the rank of  is significantly higher than the rank of , this indicates that the new f4-statistics contain information about shared drift which was not already present in the old f4-statistics, and that gene flow has occurred between R and T which is not captured by any population in L. On the other hand, if the rank of  does not increase significantly, it suggests that T does not share any drift with R which is not captured by L (there are no paths from T to any population in R which do not pass through ancestors of a population in L). This implies that all shared drift between T and R can be modeled as a linear combination of shared drift between L and R. If T is admixed between source populations which form clades with populations in L, the weights of this linear combination will correspond to the relative contributions of the source populations. The weights  are normalized so that they sum to one, and they are equivalent to the left null space of the matrix  (they satisfy ), where  is a rank  approximation of .


Whereas the purpose of qpGraph is to model the ancestral relationships among a set of populations without paying special attention to a particular population, the purpose of qpAdm and qpWave is to test whether a target population T can be modeled as a mixture of a set of left populations L (proxy source populations), relative to a set of more distantly related right populations R. No assumptions are made about the relationships among the populations in L and R.
There are simpler methods which can model a population as a mixture of other populations. For example, variations of PCA or ADMIXTURE can model allele frequencies in one population as a linear combination of allele frequencies in other populations. These methods do not require a set of right populations, so why are they necessary in the qpAdm framework? The right populations serve two purposes: First, they allow us to test whether the L is sufficient to explain all variation in T, or whether there is evidence of shared drift between T and R which is not present in L. Second, they make it possible to assign the correct weights to populations in L even in cases where the populations in L have a complex demographic history. For this it is necessary that the populations in R are differentially related to the populations in L.
In most cases the sampled left populations L will not be the true source populations for T. However, the ancestry proportions of the true source populations can still be estimated accurately if each true source population forms a clade with its representative in L relative to all other modelled populations, and if the right populations are chosen such that no two left populations form a clade relative to all right populations. While the second assumption can be tested, the first assumption is untestable because it depends on potentially unobserved populations, which should be considered when interpreting qpAdm results.
The null hypothesis tested by qpAdm is that the populations in L can account for all shared drift between T and R. A significant p-value rejecting this null hypothesis indicates that there has been gene flow between T and R (in either direction) that is not accounted for by any population in L. Often the goal of qpAdm modelling is to identify a set of populations L for which the null hypothesis doesn’t fail, and to estimate the relative contribution of each of these populations to T. However, it’s important to keep in mind that a qpAdm model with a non-significant p-value is not necessarily correct. There may be insufficient power to reject an incorrect model for a variety of reasons.
If no set of populations L can be found which fully explain the variation in T relative to R (all models fail), this may hint at the existence of a yet undiscovered “ghost” population which is related to a population in R, but not represented by any of the populations in L.
qpWave uses the same framework as qpAdm to estimate the number of independent ancestry sources which have contributed variation to T.

qpAdm works by constructing a matrix of f4-statistics between populations in L and T on one side, and populations in R on the other side. If this matrix has full rank, it suggests that each population in L and T has unique shared drift with R, and so the model fails: L alone cannot explain all shared drift between R and T. 

In more detail, qpAdm first considers all f4-statistics of the form $$ f4(L_1, L_i; R_1, R_j) $$, where $$ i $$ and $$ j $$ index all populations in L and R, respectively, except for $$ L_1 $$ and $$ R_1 $$. This $$ (n_L-1) \times (n_R-1) $$ matrix $$ X $$ of f4-statistics has full rank (all rows are linearly independent), only if all populations in L are differentially related to one or more populations in R. In other words, if no two populations in L form a clade with respect to all populations in R. Otherwise, if the rank of $$ X $$ is less than $$ n_L-1 $$, it suggests that at least one row of  $$ X $$ is redundant because there are two populations in L which form a clade with respect to all populations in R., where  and  index all populations in L and R, respectively, except for  and . This  matrix  of f4-statistics has full rank (all rows are linearly independent), only if all populations in L are differentially related to one or more populations in R. In other words, if no two populations in L form a clade with respect to all populations in R. Otherwise, if the rank of  is less than , it suggests that at least one row of   is redundant because there are two populations in L which form a clade with respect to all populations in R.
Once a combination of L and R has been found so that $$ X $$ is full rank, we add T to L and compute a new matrix $$ X' $$ of dimensions $$ (n_L+1) \times n_R $$. If the rank of $$ X' $$ is equal to the rank of $$ X $$, the inclusion of T did not add any new information relative to R: The variation in T can be fully explained by L. On the other hand, if the rank of X’ increases by one relative to the rank of $$ X' $$, it means that there is shared drift between T and some population in R which is not captured by any of the pairs in L, and thus T cannot be adequately modelled as a mix of populations in L. is full rank, we add T to L and compute a new matrix  of dimensions . If the rank of  is equal to the rank of , the inclusion of T did not add any new information relative to R: The variation in T can be fully explained by L. On the other hand, if the rank of X’ increases by one relative to the rank of , it means that there is shared drift between T and some population in R which is not captured by any of the pairs in L, and thus T cannot be adequately modelled as a mix of populations in L.
To formally test how much information is added by including T in L, $$ X' $$ is compared to a low rank approximation of itself in which the rank has an upper bound of $$ n_L $$. This low rank approximation of $$ X' $$ can be written as $$ AB $$, where $$ A $$ is $$ n_L \times n_L $$, and $$ B $$ is $$ n_L \times n_R $$ and should minimize the log likelihood $$ L(A,B) = -\frac{1}{2} E'Q^{-1}E  $$. Here, $$ E $$ is a $$ n_L (n_R-1) $$ vector of residuals given by $$ E = X' - AB $$, and $$ Q $$ is the $$ n_L (n_R-1) \times n_L (n_R-1) $$ covariance matrix of all pairs of f4-statistics in $$ X' $$, estimated using block jackknife. An initial estimate of $$ AB $$ is obtained through a singular value decomposition (SVD) of $$ X' $$. If the errors in the estimated f4-statistics were all uncorrelated, this singular value decomposition (SVD) of $$ X' $$ would already yield $$ A $$ and $$ B $$ which minimizes $$ L(A, B) $$. However, since this is generally not the case, $$ A $$ and $$ B $$ are further optimized in an iterative procedure where one matrix is held constant, while the other matrix is adjusted to minimize $$ L(A, B) $$. is compared to a low rank approximation of itself in which the rank has an upper bound of . This low rank approximation of  can be written as , where  is , and  is  and should minimize the log likelihood . Here,  is a  vector of residuals given by , and  is the  covariance matrix of all pairs of f4-statistics in , estimated using block jackknife. An initial estimate of  is obtained through a singular value decomposition (SVD) of . If the errors in the estimated f4-statistics were all uncorrelated, this singular value decomposition (SVD) of  would already yield  and  which minimizes . However, since this is generally not the case,  and  are further optimized in an iterative procedure where one matrix is held constant, while the other matrix is adjusted to minimize .
This likelihood is then compared to the likelihood of the full model, where $$ X' $$ is decomposed into two matrices without constraining the rank to be $$ n_L $$. The difference of these two log likelihoods approximately follows a $$ \Chi^2 $$ distribution with one degree of freedom, and this makes it possible to test whether the populations in L fully explain the variation in T, relative to R. The more general case where rank differences can be greater than one is described in [cite haak]. is decomposed into two matrices without constraining the rank to be . The difference of these two log likelihoods approximately follows a  distribution with one degree of freedom, and this makes it possible to test whether the populations in L fully explain the variation in T, relative to R. The more general case where rank differences can be greater than one is described in [cite haak].
The admixture weights for the populations in L relative to T are given by left null space of the matrix $$ A $$, normalized so that they sum to one., normalized so that they sum to one.


## Examples




